---
title: "P1 baseline --- toy-cohort illustration of the calibration-first pipeline"
author: "PhD student (lead), Max Moldovan, Usman Iqbal"
date: "2026-05-17"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{P1 baseline --- toy-cohort illustration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
set.seed(20260517L)

# Stan toolchain detection. brms (used for the Bayesian
# regularised-horseshoe fit) requires rstan + BH (Boost C++ headers).
# On CI runners and minimal environments these may not be installed;
# we detect their availability once and gate the Bayesian chunks
# accordingly, so the rest of the vignette still renders.
brms_ok <- requireNamespace("brms",  quietly = TRUE) &&
           requireNamespace("rstan", quietly = TRUE) &&
           requireNamespace("BH",    quietly = TRUE)
```

```{r brms-unavailable-notice, eval=!brms_ok, echo = FALSE, results = "asis"}
cat(
  "> **Note.** The Stan toolchain (`brms` + `rstan` + `BH`) is not ",
  "available in this rendering environment, so the Bayesian ",
  "regularised-horseshoe fit and its calibration / decision-curve / ",
  "discrimination columns are skipped. Re-render on a host with the ",
  "Stan toolchain installed to include the Bayesian model.\n",
  sep = ""
)
```

> **Scope.** This vignette mirrors the P1 paediatric ICU mortality pipeline
> end-to-end on a small synthetic cohort (`inst/extdata/toy_cohort.rds`,
> n = 80, mortality = 10 %). Numbers shown here are illustrative; they do
> not reproduce P1 manuscript results. The full pipeline on PIC v1.1.0
> (n = 8,736) lives at
> `../../deliverables/pccm_submission_pkg_v2/manuscript_pccm.pdf`.

# What this vignette demonstrates

The calibration-first contract for `picMort` runs through four stages:

1. **Cohort.** First ICU stay per patient, age 0--18 y, in-hospital
   mortality outcome --- the frozen contract from `vignette("cohort_spec",
   package = "picMort")`. The toy cohort obeys the same schema and the
   same invariants, downsized to keep the vignette fast.
2. **Features.** T<sub>0</sub> + 24 h prediction window, demographics
   plus vitals and labs aggregated by `min` / `max` / `mean` / `last`,
   no length-of-stay, no discharge-time, no post-window leakage. On PIC
   v1.1.0 this is `build_features(cohort, pic_paths())`; here, because
   the toy cohort ships without raw `CHARTEVENTS` / `LABEVENTS`, the
   vignette synthesises a small, deterministic feature matrix with the
   same shape and column conventions so the downstream pipeline is
   exercised faithfully. The synthesis routine is local to this vignette
   --- production runs go through `build_features()`.
3. **Models.** Penalised logistic regression (elastic net), XGBoost, and
   Bayesian logistic regression with a regularised-horseshoe prior. The
   Bayesian model is the JTM clinical-functional wedge: every patient
   carries a posterior 95 % credible interval on predicted mortality.
4. **Evaluation.** Calibration suite (slope, intercept, ICI,
   calibration-in-the-large) and decision-curve analysis at 5 %, 10 %,
   20 % thresholds are the manuscript headline. Discrimination metrics
   (AUROC, AUPRC, Brier, Brier skill score) are supporting.

```{r libs, message = FALSE, warning = FALSE}
library(picMort)
library(data.table)
```

# Stage 1 --- load the toy cohort

```{r cohort}
toy_cohort_path <- system.file("extdata", "toy_cohort.rds", package = "picMort")
cohort <- readRDS(toy_cohort_path)
dim(cohort)
table(mortality = cohort$hospital_expire_flag)
```

The toy cohort obeys the same invariants as the production cohort, scaled
to a smaller envelope (50--150 rows, mortality 5--20 %).

```{r cohort-invariants}
toy_expectations <- list(
  n_min            = 50L,
  n_max            = 150L,
  mortality_rate   = c(0.05, 0.20),
  age_range_years  = c(0, 18),
  sex_levels       = c("F", "M"),
  distinct_subject = TRUE,
  no_overlap_stays = TRUE
)
assert_cohort_invariants(cohort, expected = toy_expectations)
```

# Stage 2 --- build a small illustrative feature matrix

On PIC v1.1.0 the call is
`features <- build_features(cohort, pic_paths(), window_hours = 24L)`. The
toy cohort ships without raw event tables, so this vignette assembles a
matrix of the same shape from the cohort columns plus deterministic
synthetic vitals and labs. The hard rule is identical to production:
**no LOS, no discharge-time, no post-window timestamp ever enters the
feature matrix.**

```{r synth-features}
make_toy_features <- function(cohort, seed = 20260517L) {
  set.seed(seed)
  n <- nrow(cohort)
  baseline_hr  <- 130 - 4 * pmin(cohort$age_years, 15) + stats::rnorm(n, 0, 8)
  baseline_rr  <- 40  - 1.5 * pmin(cohort$age_years, 15) + stats::rnorm(n, 0, 4)
  baseline_sbp <- 80  + 2.5 * pmin(cohort$age_years, 15) + stats::rnorm(n, 0, 10)
  baseline_spo2 <- pmin(100, 97 + stats::rnorm(n, 0, 1.5))
  baseline_lactate <- pmax(0.5, 1.4 + 0.6 * cohort$hospital_expire_flag +
                             stats::rnorm(n, 0, 0.4))

  x <- data.table::data.table(
    icustay_id          = cohort$icustay_id,
    age_months          = cohort$age_months,
    age_years           = cohort$age_years,
    sex_male            = as.integer(cohort$sex == "M"),
    is_surgical         = as.integer(cohort$is_surgical),
    primary_icd_chapter = cohort$primary_icd_chapter,
    hr_min              = baseline_hr - abs(stats::rnorm(n, 5, 2)),
    hr_max              = baseline_hr + abs(stats::rnorm(n, 8, 3)),
    hr_mean             = baseline_hr,
    hr_last             = baseline_hr + stats::rnorm(n, 0, 4),
    rr_min              = baseline_rr - abs(stats::rnorm(n, 2, 1)),
    rr_max              = baseline_rr + abs(stats::rnorm(n, 3, 1)),
    rr_mean             = baseline_rr,
    rr_last             = baseline_rr + stats::rnorm(n, 0, 2),
    sbp_min             = baseline_sbp - abs(stats::rnorm(n, 6, 2)),
    sbp_max             = baseline_sbp + abs(stats::rnorm(n, 8, 2)),
    sbp_mean            = baseline_sbp,
    sbp_last            = baseline_sbp + stats::rnorm(n, 0, 5),
    spo2_min            = pmin(100, baseline_spo2 - abs(stats::rnorm(n, 2, 1))),
    spo2_max            = pmin(100, baseline_spo2 + abs(stats::rnorm(n, 1, 0.5))),
    spo2_mean           = baseline_spo2,
    spo2_last           = pmin(100, baseline_spo2 + stats::rnorm(n, 0, 1)),
    lactate_min         = baseline_lactate - abs(stats::rnorm(n, 0.2, 0.1)),
    lactate_max         = baseline_lactate + abs(stats::rnorm(n, 0.4, 0.2)),
    lactate_last        = baseline_lactate + stats::rnorm(n, 0, 0.2)
  )

  list(
    x            = x,
    y            = as.integer(cohort$hospital_expire_flag),
    window_hours = 24L,
    feature_set  = "toy"
  )
}

features <- make_toy_features(cohort)
dim(features$x)
```

`features$x` carries the `icustay_id` join key plus demographics, vitals
and a single representative lab. Production runs ship dozens more lab
panels; the shape is otherwise identical to `build_features()` output.

# Stage 3 --- stratified train / test split

```{r split}
split <- make_train_test_split(features, prop = 0.7, seed = 20260517L)
length(split$train_idx)
length(split$test_idx)
```

# Stage 4 --- fit the three models

Run times below are intentionally aggressive to keep the vignette under
two minutes. Production runs use wider hyperparameter grids, 1,000
bootstrap replicates for CIs, and four MCMC chains with 4,000
iterations for the Bayesian fit.

```{r fit-glmnet}
fit_en <- fit_elastic_net(features, split$train_idx, seed = 20260517L)
c(alpha = fit_en$best_alpha, lambda = fit_en$best_lambda)
```

```{r fit-xgb}
fit_xgb <- fit_xgboost(features, split$train_idx, seed = 20260517L)
fit_xgb$best_nrounds
```

```{r fit-bayes, message = FALSE, warning = FALSE, results = "hide", eval=brms_ok}
fit_bh <- fit_bayes_horseshoe(
  features, split$train_idx,
  chains = 2L, iter = 500L,
  par_ratio = 0.10, adapt_delta = 0.95,
  seed = 20260517L
)
```

```{r fit-bayes-summary, eval=brms_ok}
fit_bh$chains
fit_bh$iter
```

# Stage 5 --- predict on the held-out fold

The test fold is touched once. Predictions are aligned by
`icustay_id`; the Bayesian model additionally returns a 95 % credible
interval on the predicted probability for every test-fold patient.

```{r predict}
x_test <- features$x[split$test_idx, ]
y_test <- features$y[split$test_idx]

pred_en  <- predict_mortality(fit_en,  x_test)
pred_xgb <- predict_mortality(fit_xgb, x_test)
if (brms_ok) {
  pred_bh <- predict_mortality(fit_bh, x_test)
  head(pred_bh, 3)
} else {
  pred_bh <- NULL
}
```

# Stage 6 --- evaluation (the manuscript headline)

```{r eval, warning = FALSE}
probs <- list(
  elastic_net = pred_en$prob_raw,
  xgboost     = pred_xgb$prob_raw
)
if (brms_ok) {
  probs$bayes_horseshoe <- pred_bh$prob_raw
}

calib <- lapply(probs, calibration_suite,
                y = y_test, n_boot = 200L, seed = 20260517L)
dca   <- decision_curve(probs, y_test,
                        thresholds = c(0.05, 0.10, 0.20),
                        plot_grid = FALSE)
disc  <- discrimination_metrics(probs, y_test,
                                reference = "elastic_net",
                                n_boot = 200L, seed = 20260517L)
```

## Calibration headline

```{r calib-table}
calib_tbl <- data.table::rbindlist(lapply(names(calib), function(nm) {
  data.table::data.table(
    model     = nm,
    slope     = calib[[nm]]$slope[["estimate"]],
    intercept = calib[[nm]]$intercept[["estimate"]],
    ici       = calib[[nm]]$ici[["estimate"]],
    cit_large = calib[[nm]]$cit_large[["estimate"]]
  )
}))
knitr::kable(calib_tbl, digits = 3,
             caption = "Calibration on the toy-cohort held-out fold.")
```

## Decision-curve net benefit at clinically relevant thresholds

```{r dca-table}
dca_headline <- dca[threshold %in% c(0.05, 0.10, 0.20) & type == "model"]
knitr::kable(dca_headline, digits = 4,
             caption = "Net benefit at 5%, 10%, 20% thresholds.")
```

## Discrimination (supporting)

```{r disc-table}
knitr::kable(disc, digits = 3,
             caption = "Discrimination metrics with bootstrap 95% CIs.")
```

# How this maps to the P1 manuscript

This vignette walks the same four-stage pipeline that drives the P1
*Pediatric Critical Care Medicine* manuscript bundle at
`deliverables/pccm_submission_pkg_v2/manuscript_pccm.pdf`. The
production bundle:

- runs on PIC v1.1.0 (n = 8,736, mortality 0.0844) rather than the 80-row
  toy cohort,
- uses the full feature panel from `build_features()` (vitals + labs from
  `CHARTEVENTS` and `LABEVENTS`, 13 panels) under the same T<sub>0</sub>
  + 24 h prediction window,
- adds the PIM3 baseline reconstructed via `compute_pim3()` (Gate G3),
- runs the Bayesian fit with four chains and 4,000 iterations, with
  patient-level 95 % credible intervals as the clinical-functional
  headline,
- reports calibration and decision-curve analysis as the primary
  outcome, with discrimination supporting.

Numbers in the tables above are not stable across runs of the toy
synthesiser and must not be quoted in any paper. The toy cohort exists
to let `R CMD check` exercise the package surface end-to-end and to give
new contributors a runnable scaffold.

```{r session}
sessionInfo()
```
