---
title: "P2 inference --- globally-optimal CIs for clinical predicted probabilities (scaffold)"
author: "Max Moldovan (lead), Stats PhD student, Usman Iqbal"
date: "2026-05-17"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{P2 inference --- inductive-CI scaffold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
## Vignette-wide: do not execute illustrative chunks at build time.
knitr::opts_chunk$set(eval=FALSE)
```

> **Status --- placeholder.** This vignette is a pre-publication scaffold
> for P2 of the picMort series. The methodological draft is not yet
> written; this file describes the contribution, points to the P2-ready
> artefacts produced by P1, and sketches the form of the inductive-CI
> machinery so a future session can pick the work up cleanly. No code
> chunk is executed at vignette-build time (chunks are tagged with
> `eval=FALSE` in the setup block above). The vignette will be replaced
> with a runnable end-to-end illustration once the methodology lands.

# Methodological contribution

P2 applies Max Moldovan's globally-optimal / inductive confidence-interval
framework to inference on **predicted probabilities** from a clinical
prediction model, and benchmarks it against the Bayesian credible
intervals delivered by the regularised-horseshoe fit in P1.

The wedge is finite-sample honesty. Standard practice for predicted-
probability uncertainty in clinical machine learning is one of (i)
bootstrap percentile intervals, which under-cover at small *n* and on
extreme probabilities; (ii) Bayesian credible intervals, which inherit
the prior and the model class; or (iii) no intervals at all. The
inductive framework constructs a confidence region that is optimal under
the worst-case data-generating distribution in a specified neighbourhood
of the empirical distribution, with finite-sample coverage guarantees
that do not depend on asymptotics or prior choice.

The target venue is *Statistics in Medicine* (methods lead), with a
clinical mirror submission to a paediatric or critical-care journal once
the method paper is in press.

# Inheritance from P1

P2 is gated on P1's calibration and decision-curve results landing. The
cohort and feature pipeline are inherited verbatim from
`vignette("cohort_spec", package = "picMort")` and the manuscript at
`../../deliverables/pccm_submission_pkg_v2/manuscript_pccm.pdf`. The
saved P1 artefacts that P2 will consume:

- `../../scripts/_outputs/fits_g4.rds` --- the three fitted models from
  P1 Gate G4 (elastic net, XGBoost, Bayesian horseshoe).
- `../../scripts/_outputs/evaluation_g5_g6.rds` --- the evaluation suite
  from Gates G5 (calibration + DCA) and G6 (discrimination + subgroups),
  including the predicted-probability vectors on the held-out fold.

These pointers are documented for the next session; the files are
*not* loaded at vignette-build time. They live outside the package
source tree and may not be present in the build sandbox or the r-universe
binary.

# Form of the inductive-CI machinery

The shape of the API is sketched below for orientation. The
implementation will live in a new `R/inductive_ci.R` module once the
mathematics has been written up and externally reviewed.

```{r form-sketch}
#' Globally-optimal CI for a predicted probability
#'
#' Constructs a confidence region for the conditional probability
#' P(Y = 1 | X = x_*) under finite-sample coverage in the
#' inductive-conformal sense.
#'
#' @param model A fitted model object understood by `predict_mortality()`.
#' @param new_data New `data.table` of test-fold patients (same schema as
#'   `features$x` in P1).
#' @param calibration Held-out calibration fold (a list with `x`, `y` of
#'   the same shape as P1's `features`). Required for the inductive
#'   construction.
#' @param level Nominal coverage level. Default `0.95`.
#'
#' @return A `data.table` aligned to `nrow(new_data)` with columns
#'   `prob`, `lower`, `upper`, `width`, and `coverage_basis`
#'   (`"inductive"` or `"asymptotic_fallback"`).
inductive_ci <- function(model, new_data, calibration, level = 0.95) {
  ## Implementation deferred to P2 G2.
  ## Will share the prep / bake plumbing used in predict_mortality() so
  ## the same fitted model can carry either the Bayesian credible
  ## interval (from P1) or the inductive interval (from P2) without
  ## refitting.
  NULL
}
```

# Open methodological items

1. **Conformal score choice.** Default candidates: log-loss residual,
   Brier residual, and a calibration-residual score. The choice
   interacts with whether the inductive band reduces neatly to the
   patient-level Bayesian CrI under exchangeability assumptions.
2. **Stratified vs marginal coverage.** Clinical use typically wants
   per-subgroup coverage (age strata, surgical / medical, primary ICD
   chapter). The inductive construction admits stratified coverage
   directly; the benchmarking section needs to compare against
   stratified bootstrap and stratified Bayesian credible intervals on
   the same partition.
3. **Comparison metric.** The natural pair is (mean width, empirical
   coverage). Add a *width-vs-coverage* frontier across nominal levels
   `c(0.80, 0.90, 0.95, 0.99)` to expose the regime where the inductive
   construction wins materially.

# Next concrete step

When P2 reopens, the smallest action is: load `fits_g4.rds` and
`evaluation_g5_g6.rds` from the P1 outputs, extract the held-out fold's
predicted probabilities and the calibration partition, and prototype
`inductive_ci()` against the Bayesian CrIs already in
`pred_bh$prob_lower` / `pred_bh$prob_upper`. P1 reports a mean Bayesian
CrI width of 0.083 on the held-out fold; the inductive construction
needs to beat that envelope at matched coverage to clear the
methodological gate.
