---
title: "P3 fusion --- two-tower structured + Chinese clinical-note model (scaffold)"
author: "Computational PhD student (lead), Max Moldovan, Usman Iqbal"
date: "2026-05-17"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{P3 fusion --- two-tower scaffold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
## Vignette-wide: do not execute illustrative chunks at build time.
knitr::opts_chunk$set(eval=FALSE)
```

> **Status --- placeholder.** This vignette is a pre-publication scaffold
> for P3 of the picMort series. The methodological work is gated on two
> independent items: (a) a one-day audit of the PIC v1.1.0
> `EMR_SYMPTOMS` table to confirm content type and coverage, and (b)
> clarification of paper attribution given that an external collaborator
> already holds an LLM workstream on free-text notes (see
> `usman_feedback_scope_gap_2026-05-11`). The vignette will be replaced
> with a runnable two-tower illustration once both items resolve. No
> code chunk is executed at vignette-build time.

# Methodological contribution

P3 evaluates whether Chinese clinical-note embeddings from PIC v1.1.0's
`EMR_SYMPTOMS` table add discriminative or calibration signal **beyond**
the structured first-24-hour feature panel that P1 establishes. The
proposed architecture is a two-tower model:

- **Structured tower.** The P1 baseline pipeline frozen at the
  best-calibrated model from P1's evaluation (likely the Bayesian
  regularised-horseshoe fit on calibration grounds, pending P1 G6).
- **Text tower.** Token-level or sentence-level embeddings of
  `EMR_SYMPTOMS` entries dated $\le T_0 + 24$ h, encoded by a
  pre-trained Chinese clinical language model (candidate: MC-BERT or a
  Chinese biomedical BERT variant; the exact backbone is locked at P3
  Gate G2).
- **Fusion head.** A small calibration-aware fusion layer (concatenation
  + low-rank interaction + temperature scaling) that produces a single
  probability of in-hospital mortality, with patient-level uncertainty
  retained from the structured tower's Bayesian fit.

The reporting standard is calibration-first, the same contract as P1:
calibration slope / intercept / ICI plus decision-curve net benefit at
5 %, 10 %, 20 % thresholds are the headline. Discrimination metrics
serve a supporting role. The two-tower fusion claim must clear *both*
calibration and net-benefit gates relative to the structured-only
baseline --- a discrimination-only gain is not the finding.

The target venue is *npj Digital Medicine* or *Lancet Digital Health*
(clinical machine learning lead), with *JAMIA* as a secondary route.

# Open gating items

## (a) `EMR_SYMPTOMS` audit

Before any text tower is trained, PIC v1.1.0's `EMR_SYMPTOMS` table needs
a one-day audit answering four questions:

1. **Language.** Is the text genuinely free-form Chinese clinical
   narrative, or is it template-filled / structured-disguised-as-text?
   The vignette working title in `vignettes/paper3_fusion.Rmd` assumes
   the former; the latter demotes the contribution materially.
2. **Structure.** Free text vs structured codes vs hybrid. Determines
   whether a language-model embedding even applies as the encoding
   choice.
3. **Coverage.** What fraction of the cohort has any `EMR_SYMPTOMS`
   entry dated within $[T_0, T_0 + 24)$ h? If coverage is sparse, the
   fusion claim becomes a missingness-handling claim instead.
4. **Temporal hygiene.** No post-window entries leak into the text
   tower. The same `audit_no_leakage()` machinery from P1 extends to
   the text-event provenance once a per-event timestamp column is
   confirmed.

## (b) Paper attribution

A note in the series' working memory
(`usman_feedback_scope_gap_2026-05-11`) records that an external
collaborator already holds an LLM workstream on free-text clinical
notes. The P3 paper-attribution is therefore **open**: the picMort
team's P3 may fold into that workstream as co-authorship, audit the
collaborator's pipeline against the calibration-first contract, or
remain an independent contribution depending on what has already been
delivered. The next session must ask before scoping P3 further.

# Pre-audit sketch of the structured tower API

The structured tower reuses P1 verbatim. The fusion API will live in a
new `R/fusion.R` module once the audit confirms the text-tower
direction.

```{r form-sketch}
#' Two-tower fusion of structured + text predictions
#'
#' Combines a structured-tower fit (any object from `fit_elastic_net()`,
#' `fit_xgboost()`, or `fit_bayes_horseshoe()`) with a frozen text-tower
#' embedding model into a single calibrated mortality prediction.
#'
#' @param structured_model A fitted structured model from P1.
#' @param text_embeddings A `data.table` keyed on `icustay_id` carrying
#'   pre-computed text-tower embeddings (one row per cohort patient,
#'   one column per embedding dimension).
#' @param fusion One of `"concat"` (default --- linear combination on
#'   the logit scale), `"low_rank"` (rank-r bilinear interaction), or
#'   `"calibration_only"` (text features used only for temperature
#'   scaling, never for a separate score). The third option is the
#'   pre-registered null fusion against which the first two must clear
#'   a calibration + DCA gate.
#'
#' @return An object compatible with `predict_mortality()`.
fit_two_tower <- function(structured_model, text_embeddings,
                          fusion = c("concat", "low_rank", "calibration_only")) {
  ## Implementation deferred to P3 G3 (post-audit).
  NULL
}
```

# Next concrete step

When P3 reopens, the smallest action is to land the `EMR_SYMPTOMS`
audit. The audit script does not need to live inside `picMort` ---
it inherits the cohort builder via `pic_paths()$emr_symptoms` and
produces a short audit report. Until the audit answers the four
questions above and the collaborator attribution is resolved, this
vignette is the correct shape: a placeholder that documents the
contribution and the gates rather than running code that would mislead
about what has been done.
